Erosion of somatic tissue identity with loss of the X-linked intellectual disability factor KDM5C

### 2024-09-16 submission/manuscript.Rmd
res 2 edits

We then evlauted testicular DEG expression in wild-type testes versus testes with germ cell depletion[@muellerIndependentSpecializationHuman2013], which was accomplished by heterozygous _W_ and _Wv_ mutations in the enzymatic domain of _c-Kit_ (Kit^W/Wv^)[@handelSertoliCellDifferentiation1979]. Almost all _Kdm5c_-KO testis-enriched DEGs lost expression with germ cell depletion (Figure 2B). We then assessed testis-enriched DEG expression in a published single cell RNA-seq dataset that identified cell type-specific markers within the testis[@greenComprehensiveRoadmapMurine2018]. Some _Kdm5c_-KO testis-enriched DEGs were classified as specific markers for different germ cell developmental stages (e.g. spermatogonia, spermatocytes, round spermatids, and elongating spermatids), yet none marked somatic cells (Figure 2C). Together, these data demonstrate that the _Kdm5c_-KO brain aberrantly expresses germline genes, but not somatic testicular genes, reflecting an erosion of the soma-germline boundary.

As of yet, research on germline gene silencing mechanisms has focused on a handful of key genes rather than assessing germline gene suppression genome-wide due to the lack of a comprehensive gene list. We therefore generated a list of mouse germline-enriched genes using RNA-seq datasets of Kit^W/Wv^ mice that included males and females at embryonic day 12, 14, and 16[@sohGeneRegulatoryProgram2015] and adult male testes[@muellerIndependentSpecializationHuman2013]. We defined genes as germline-enriched if their expression met the following criteria: 1) their expression is greater than 1 FPKM in wild-type gonads 2) their expression in any non-gonadal tissue of adult wild type mice[@liComprehensiveMouseTranscriptomic2017] does not exceed 20% of their maximum expression in the wild-type germline, and 3) their expression in the germ cell-depleted gonads, for any sex or time point, does not exceed 20% of their maximum expression in the wild-type germline. These criteria yielded `r nrow(germgenes)` germline-enriched genes (Figure 2D), which was hereafter used as a resource to globally characterize germline gene misexpression with _Kdm5c_ loss (Supplementary table 1).

>>>>

We then evlauted testicular DEG expression in wild-type testes versus testes with germ cell depletion[@muellerIndependentSpecializationHuman2013], which was accomplished by heterozygous _W_ and _Wv_ mutations in the enzymatic domain of _c-Kit_ (Kit^W/Wv^)[@handelSertoliCellDifferentiation1979]. Almost all _Kdm5c_-KO testis-enriched DEGs lost expression with germ cell depletion (Figure 2B). We then assessed testis-enriched DEG expression in a published single cell RNA-seq dataset that identified cell type-specific markers within the testis[@greenComprehensiveRoadmapMurine2018]. Some _Kdm5c_-KO testis-enriched DEGs were classified as specific markers for different germ cell developmental stages (e.g. spermatogonia, spermatocytes, round spermatids, and elongating spermatids), yet none marked somatic cells (Figure 2C). Together, these data demonstrate that the _Kdm5c_-KO brain aberrantly expresses germline genes but not somatic testicular genes, reflecting an erosion of the soma-germline boundary.

As of yet, research on germline gene silencing mechanisms has focused on a handful of key genes rather than assessing germline gene suppression genome-wide, due to the lack of a comprehensive gene list. We therefore generated a list of mouse germline-enriched genes using RNA-seq datasets of Kit^W/Wv^ mice that included males and females at embryonic day 12, 14, and 16[@sohGeneRegulatoryProgram2015] and adult male testes[@muellerIndependentSpecializationHuman2013]. We defined genes as germline-enriched if their expression met the following criteria: 1) their expression is greater than 1 FPKM in wild-type gonads 2) their expression in any non-gonadal tissue of adult wild type mice[@liComprehensiveMouseTranscriptomic2017] does not exceed 20% of their maximum expression in the wild-type germline, and 3) their expression in the germ cell-depleted gonads, for any sex or time point, does not exceed 20% of their maximum expression in the wild-type germline. These criteria yielded `r nrow(germgenes)` germline-enriched genes (Figure 2D), which was hereafter used as a resource to globally characterize germline gene misexpression with _Kdm5c_ loss (Supplementary table 1).
